Interactive Curvature Tensor Visualization on Digital Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow} ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex} ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}

### 2016-01-19 latex/article.tex
Passe 1 sur le DGCI (désolée David j'ai tout pété la section 4 ... tu lis, tu me dis ce que tu en penses et si ça va pas je reviendrai à l'ancienne version)

Volumetric objects are being more and more popular in many
applications ranging from object modeling and rendering in Computer
Graphics to geometry processing in Medical Imaging or Material
Sciences. When considering large volumetric data, interactive
%visualization of such images (or isovalue surfaces) is a complex
visualization of those objects (or isosurfaces) is a complex
problem. Such issues become even more difficult when dynamic
%volumetric datasets are considered. Beside such isosurface
volumetric datasets are considered. Beside
visualization, we are also interested in performing geometry
processing on the digital object and to explore different parameter
settings of the geometry processing tool.  In this work, we focus on
interactive visualization of volumetric data isosurfaces combined with
a curvature tensor (mean/Gaussian curvature, principal curvatures
directions, \ldots) estimation at each point of the digital
surface.
%Curvature maps on a given isosurface are key information when
%exploring 3D structures in Material Sciences \todo{REF}.
Furthermore,
most curvature estimators require a parameter fixing the scale
at which the computation is performed. For short, such parameter
(integration radius, convolution kernel size, \ldots) specifies a scale for analyzing the object surface, and is naturally related to the amount of perturbation on input data. As a consequence, exploring interactively
different values of such parameter is mandatory.
\vspace{0.2cm}

@@

@@

In \cite{DGCI2013}, estimators are defined based on the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
Estimators of both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information are proved to be multigrid convergent
property \cite{DGCI2013}. For short, when the digital object becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

In \cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
These estimators are proven convergent when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}. More formally, it leads to:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

@@

We first present design principles for the GPU implementation and then
present our approaches.

\subsection{Design principles}

The general idea is to perform an Integral Invariant computation on
GPU at each vertex of the generated triangulated mesh. First note that
when the LoD criterion is removed, each MC vertex is exactly centered at
a surfel center. At different depth of the octree, MC vertices still
correspond to surfel centers of subsampled versions of the input
object. Hence, Integral Invariant framework defined in
Section~\ref{sec:curv-tens-estim} is consistent with the triangulated
surface obtained on GPU: triangles are used for visualization purposes
but all computations are performed on the digital object $\DSh$ for
estimators
(\ref{eq:dig-curvature-estimator-k}),
(\ref{eq:dig-curvature-estimator-H}), (\ref{eq:dig-curvature-estimator-k1})
and (\ref{eq:dig-curvature-estimator-k2}).

To implement the integration on $\Ball{R}{\vx} \cap \Shape$
(Fig.~\ref{fig:approx}-$a$), a naive approach consists in scanning all digital
points inside such domain (Fig.~\ref{fig:approx}-$b$) and then to estimate the
geometrical moment as the sum of the geometrical moments of each elementary
cubes lying in the intersection. On GPU, \emph{mipmap} textures are used to
obtain a multi-resolution information: if the input binary object is stored in a
3D texture, GPU hardware can construct a multi-resolution pyramid such that 8
neighboring voxel intensities at level $l$ are averaged to define the voxel
value at level $l+1$. If the level 0 corresponds to the binary input object, at
a given level $l$, a voxel $(x,y,z)$ contains the fraction of $\DSh$ belonging
to the cube of center $(x,y,z)$ and edge length $2^l$. As a consequence, we can
approximate the volume of $\Ball{R}{\vx}\cap\Shape$ by considering mipmap values
at a given resolution $l$ (Fig.~\ref{fig:approx}-$c$). In this case, errors only
occurs for cells lying inside $\Shape$ (with density 1) not entirely covered by
$\Ball{R}{\vx}$. Furthermore, the \emph{mipmap} texture can be used to design,
using a single texture probe, a fast inclusion test of a given cell $c$ at level
$l$ into the shapes: we say that $c$ is in $\Shape$ if its density (retrieved
from the texture probe at level l) is greater than $1/2$. The idea here is to
mimic a kind of adaptive Gauss digitization process.

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=4.0cm]{figs/approx1}}
    {\includegraphics[width=4.0cm]{figs/approx-reg-1}}
    {\includegraphics[width=4.0cm]{figs/approx-reg-2}}
  \end{center}
  \vspace{-0.35cm}
  \caption{Integral computation with an area estimation (\emph{left}) from pixel
    counting (\emph{middle}) or from densities (represented by different
    gray values) in a mipmap approximation of $\Shape$ (\emph{right}).}
  \label{fig:approx}
  \vspace{-0.35cm}
\end{figure}

Using the hierarchical nature of the mipmap texture, we could also
consider hierarchical decompositions of either $\Ball{R}{\vx}$
(Fig.~\ref{fig:approx2}-$a$) or $\Ball{R}{\vx}\cap\Shape$
(Fig.~\ref{fig:approx2}-$b$). The idea would be to decompose the
regions into mipmap cells of different resolution in order to limit
the number of texture access (important bottleneck on GPU hardwares)
and to get better approximated quantities. Note that the scenario in
Fig.~\ref{fig:approx2}-$b$ is clearly the best approach: the
volume is exact  (\wrt the regular grid case) since geometrical moments
are additive quantities, and the number of texture fetches is  optimal
at this point.

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=4cm]{figs/approx-bh}}
    {\includegraphics[width=4cm]{figs/approx-ada}}
  \end{center}
  \vspace{-0.35cm}
  \caption{Adaptive approximation of the integration using a
    hierarchical decomposition of the ball $\Ball{R}{\vx}$ (\emph{left}), or a decomposition of $\Ball{R}{\vx}\cap\Shape$ (\emph{right}).}
  \label{fig:approx2}
  \vspace{-0.35cm}
\end{figure}

@@

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on the GPU %hardware,
this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierarchical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU %mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells in $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).

\paragraph{Dynamic Refinement.} In this approach, we want to optimize
the interactivity and the execution flow or parallelism on GPU. The
integration is simply computed using a regular grid at different
resolution $l$ (Fig. \ref{fig:approx}-$b$ and $c$). The shader code
becomes trivial (simple triple loop) and a lot of texture fetches are
involved, but interactivity can be easily obtained. Indeed, we
consider the following multi-pass approach:
\begin{enumerate}
\item When the surface geometry has been computed, we set the current
  level $l$ at an high level $l\EqDef l_{max}$ of the mipmap
  pyramid.
\item We compute the integrals (and the curvature tensor) at the
  mipmap level $l$ and  send the estimated quantities to the
  visualization stage.
\item If the user changes the camera position or the computation
  parameters, we return to step 1.
\item If the current framerate is above a given threshold, we decrease
  $l$ and return to step 2 if the final $l_0$ level has not been
  reached yet.
\end{enumerate}
A consequence of the multi-pass approach is that when there is no
interaction (camera settings, parameters), the GPU automatically
refines the estimation. Even if step 2 is quite expensive, in
$O\left(\left(\frac{R}{2^l}\right)^3\right)$, the interactive control
 in step 3 considerably improves the interactive exploration of the
 tensor with fast preliminary approximations which are quickly refined.

@@

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=2.2cm]{figs/Octa_mean}}
    {\includegraphics[width=2.2cm]{figs/Octa_gaussian}}
%    {\includegraphics[width=2.7cm]{figs/Octa_k1}}
%    {\includegraphics[width=2.7cm]{figs/Octa_k2}}\\
%    {\includegraphics[width=2.7cm]{figs/Octa_dir_min}}
%    {\includegraphics[width=2.7cm]{figs/Octa_dir_max}}
    {\includegraphics[height=2.0cm]{figs/Octa_dir_min_zoom}}
    {\includegraphics[height=2.0cm]{figs/Octa_dir_max_zoom}}\\
    % {\includegraphics[height=4.0cm]{figs/EG_dragon_normal_zoom}}
    {\includegraphics[width=2.2cm]{figs/function_mean_0}}
    {\includegraphics[width=2.2cm]{figs/function_mean_1}}
    {\includegraphics[width=2.2cm]{figs/function_mean_2}}
    {\includegraphics[width=2.2cm]{figs/function_mean_3}}
    {\includegraphics[width=2.2cm]{figs/function_mean_4}}
  \end{center}
  \vspace{-0.35cm}
  \caption{\emph{First row}: Mean and Gaussian curvature estimation, (zoom of) first and second principal directions estimation on ``OctaFlower'' with a digital domain of $130^3$.
  % \emph{Second row:} (Zoom of) First and second principal directions estimation on an OctaFlower.
  \emph{Second row:} Mean curvature computed in real-time on a dynamic object.}
  \label{fig:full}
  \vspace{-0.35cm}
\end{figure}

>>>>

Volumetric objects are being more and more popular in many
applications ranging from object modeling and rendering in Computer
Graphics to geometry processing in Medical Imaging or Material
Sciences. When considering large volumetric data, interactive
%visualization of such images (or isovalue surfaces) is a complex
visualization of those objects (or isosurfaces) is a complex
problem. Such issues become even more difficult when dynamic
%volumetric datasets are considered. Beside such isosurface
volumetric datasets are considered. Beside
visualization, we are also interested in performing geometry
processing on the digital object and to explore different parameter
settings of the geometry processing tool.
\color{blue}
Here, we focus on curvature tensor estimators.
Most of them require a parameter fixing the scale
at which the computation is performed. For short, such parameter
(integration radius, convolution kernel size, \ldots) specifies a scale for analyzing the object surface, 
and is naturally related to the amount of perturbation on input data. As a consequence, when using
such tensors, exploring different values of this parameter is mandatory.
However, this is usually an offline process, due to the amount of computations that
needs to be done.
\normalcolor
%In this work, we present an interactive visualization of volumetric data isosurfaces combined with
%a curvature tensor (mean/Gaussian curvature, principal curvatures
%directions, \ldots) estimation at each point of the digital
%surface.
%Curvature maps on a given isosurface are key information when
%exploring 3D structures in Material Sciences \todo{REF}.

@@

\vspace{0.2cm}
\color{blue}
In this paper, we will first present the previous works on
curvature tensor estimation. Then, we will see how we can
efficiently extract a triangulation of an object on the GPU
and finally, we will present how we can evaluate the curvature tensor in
real time and discuss on the several approaches that we tried.
\normalcolor

@@

In \cite{DGCI2013}, estimators are defined based on the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
Estimators of both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information are proved to be multigrid convergent
property \cite{DGCI2013}. For short, when the digital object becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
\color{blue}
associated one on $\Bd{\Shape}$ in $O\left(h^{\frac{1}{3}}\right)$ for
convex shapes with at least $C^3$-boundary and bounded curvature.
\normalcolor
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
%Setting $R = kh^\frac{1}{3}$ on convex shapes
%with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
%related to the maximal curvature of the shape, they have:
%%
%\begin{align}
%\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
%\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
%&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
%&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
%\end{align}

In \cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
\color{blue}
These estimators are proven convergent in $O\left(h^{\frac{1}{3}}\right)$ when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}.
\normalcolor 
%More formally, it leads to:
%%
%\begin{align}
%  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
%  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
%  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
%  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
%\end{align}

@@

We first present design principles for the GPU implementation and then
present our approaches.
%\subsection{Design principles}
The general idea is to perform an Integral Invariant computation on
GPU at each vertex of the generated triangulated mesh.
\color{blue}
Since GPU's are massively parallel, we can do all those computations in parallel 
to get a very efficient implementation. Please note that
\normalcolor
when the LoD criterion is removed, each MC vertex is exactly centered at
a surfel center. At different depth of the octree, MC vertices still
correspond to surfel centers of subsampled versions of the input
object. Hence, Integral Invariant framework defined in
Section~\ref{sec:curv-tens-estim} is consistent with the triangulated
surface obtained on GPU: triangles are used for visualization purposes
but all computations are performed on the digital object $\DSh$ for
estimators
(\ref{eq:dig-curvature-estimator-k}),
(\ref{eq:dig-curvature-estimator-H}), (\ref{eq:dig-curvature-estimator-k1})
and (\ref{eq:dig-curvature-estimator-k2}).

\color{blue}
To implement the integration on $\Ball{R}{\vx} \cap \Shape$ (Fig.~\ref{fig:approx}-$a$), we tried several approaches.

@@

\paragraph{Naive Approach.}
A first simple solution , consists in scanning all digital
\normalcolor
points inside the integration domain (Fig.~\ref{fig:approx}-$b$) and then to estimate the
geometrical moment as the sum of the geometrical moments of each elementary
cubes lying in the intersection. 
\color{blue}
This is exactly similar to what is done on the CPU, but on the GPU, we can
exploit \emph{mipmap} textures to obtain a multi-resolution information.
\normalcolor
If the input binary object is stored in a
3D texture, GPU hardware constructs a multi-resolution pyramid (mipmap) such that 8
neighboring voxel intensities at level $l$ are averaged to define the voxel
value at level $l+1$. If the level 0 corresponds to the binary input object, at
\color{blue}
a given level $l$, a texture probe returns a voxel $(x,y,z)$, containing the fraction of $\DSh$ belonging
\normalcolor
to the cube of center $(x,y,z)$ and edge length $2^l$. As a consequence, we can
approximate the volume of $\Ball{R}{\vx}\cap\Shape$ by considering mipmap values
at a given resolution $l$ (Fig.~\ref{fig:approx}-$c,d$). In this case, errors only
occurs for cells lying inside $\Shape$ (with density 1) not entirely covered by
$\Ball{R}{\vx}$. Furthermore, the \emph{mipmap} texture can be used to design,
using a single texture probe, a fast inclusion test of a given cell $c$ at level
$l$ into the shapes: we say that $c$ is in $\Shape$ if its density (retrieved
from the texture probe at level l) is greater than $1/2$. The idea here is to
mimic a kind of adaptive Gauss digitization process.

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=4.0cm]{figs/approx1}}
    {\includegraphics[width=4.0cm]{figs/approx-reg-1}}
    {\includegraphics[width=4.0cm]{figs/approx-reg-2}}
    %TODO image Mipmap 3eme Niveau
  \end{center}
  \vspace{-0.35cm}
  \caption{Integral computation with an area estimation (\emph{left}) 
  \color{blue}
  	from mipmaps at level $l=0$ (note that this is the same as pixel counting) \emph{middle-left}
  	from mipmaps at level $l=1$ \emph{middle-right}
  	from mipmaps at level $l=2$ \emph{right}
  	We can note that the higher the mipmap level, the more we approximate our integrand, and the less texture probe we need to do.
    \normalcolor}
  \label{fig:approx}
  \vspace{-0.35cm}
\end{figure}

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=4cm]{figs/approx-bh}}
    %{\includegraphics[width=4cm]{figs/approx-ada}}
    %TODO Image alignement cellules mipmaps
  \end{center}
  \vspace{-0.35cm}
  \caption{Adaptive approximation of the integration using a
    hierarchical decomposition of the ball $\Ball{R}{\vx}$ }
    %(\emph{left}), or a decomposition of $\Ball{R}{\vx}\cap\Shape$ (\emph{right}).}
  \label{fig:approx2}
  \vspace{-0.35cm}
\end{figure}

@@

\paragraph{Hierarchical Decomposition.}  
\color{blue}
Using the hierarchical nature of the mipmap texture, we could also
consider hierarchical decompositions of $\Ball{R}{\vx}$
(Fig.~\ref{fig:approx2}-$a$) 
%or $\Ball{R}{\vx}\cap\Shape$
%(Fig.~\ref{fig:approx2}-$b$). 
The idea would be to decompose the
regions into mipmap cells of different resolution in order to limit
the number of texture access (important bottleneck on GPU hardwares)
and to get better approximated quantities. 
%Note that the scenario in
%Fig.~\ref{fig:approx2}-$b$ is clearly the best approach: the
%volume is exact  (\wrt the regular grid case) since geometrical moments
%are additive quantities, and the number of texture fetches is  optimal
%at this point.
%As discussed above, best
%results in terms of texture fetch can be obtained by the approach
%presented in Fig.~\ref{fig:approx2}-$b$. 
%At each vertex, the
%computation algorithm would consist in a recursive decomposition of
%the domain such that a cell is subdivided if it does not belong to
%$\Ball{R}{\vx}\cap\Shape$. However, on the GPU %hardware,
%this approach is
%highly inefficient since the hardware optimizes the parallelism only
%when the micro-program described in the shader has a predictive
%execution flow. Hence, implementing the recursive algorithm (or its
%de-recursified version) requires a lot of branches (conditional
%\emph{if} tests) in the flow. 
We have thus considered a fast multi-resolution approximation as
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierarchical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU %mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells in $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).
Still, this method is not fully exploiting the GPU parallelism. 
The hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing this recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. 
\normalcolor

\paragraph{Dynamic Refinement.} In this last approach, we want to optimize
the interactivity and the execution flow or parallelism on GPU. The
integration is simply computed using a regular grid at different
resolution $l$ (Fig. \ref{fig:approx}-$b$ and $c$). The shader code
becomes trivial (simple triple loop) and a lot of texture fetches are
involved, but interactivity can be easily obtained. Indeed, we
consider the following multi-pass approach:
\begin{enumerate}
\item When the surface geometry has been computed, we set the current
  level $l$ at an high level $l\EqDef l_{max}$ of the mipmap
  pyramid.
\item We compute the integrals (and the curvature tensor) at the
  mipmap level $l$ and  send the estimated quantities to the
  visualization stage.
\item If the user changes the camera position or the computation
  parameters, we return to step 1.
\item If the current framerate is above a given threshold, we decrease
  $l$ and return to step 2 if the final $l_0$ level has not been
  reached yet.
\end{enumerate}
A consequence of the multi-pass approach is that when there is no
interaction (camera settings, parameters), the GPU automatically
refines the estimation. Even if step 2 is quite expensive, in
$O\left(\left(\frac{R}{2^l}\right)^3\right)$, the interactive control
in step 3 considerably improves the interactive exploration of the
tensor with fast preliminary approximations which are quickly refined.

@@

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=2.2cm]{figs/Octa_mean}}
    {\includegraphics[width=2.2cm]{figs/Octa_gaussian}}
%    {\includegraphics[width=2.7cm]{figs/Octa_k1}}
%    {\includegraphics[width=2.7cm]{figs/Octa_k2}}\\
%    {\includegraphics[width=2.7cm]{figs/Octa_dir_min}}
%    {\includegraphics[width=2.7cm]{figs/Octa_dir_max}}
    {\includegraphics[height=2.0cm]{figs/Octa_dir_min_zoom}}
    {\includegraphics[height=2.0cm]{figs/Octa_dir_max_zoom}}\\
    % {\includegraphics[height=4.0cm]{figs/EG_dragon_normal_zoom}}
    {\includegraphics[width=2.2cm]{figs/function_mean_0}}
    {\includegraphics[width=2.2cm]{figs/function_mean_1}}
    {\includegraphics[width=2.2cm]{figs/function_mean_2}}
    {\includegraphics[width=2.2cm]{figs/function_mean_3}}
    {\includegraphics[width=2.2cm]{figs/function_mean_4}}
  \end{center}
  \vspace{-0.35cm}
  \caption{\emph{First row}: Mean and Gaussian curvature estimation, (zoom of) first and second principal directions estimation on ``OctaFlower'' with a digital domain of $130^3$.
  % \emph{Second row:} (Zoom of) First and second principal directions estimation on an OctaFlower.
  \emph{Second row:} Mean curvature computed in real-time (around 20 FPS) on a dynamic object.}
  \label{fig:full}
  \vspace{-0.35cm}
\end{figure}
